---
title: "L-1.1. Applications leading to differential equations"
subtitle: "`MAT211`. DiffEq. Lecture 1."
author: "E. Valderrama-Araya, Ph.D.  |  <evalderrama@oru.edu>"
format: 
  revealjs:
    theme: default #beige blood dark default league moon night serif simple sky solarized
    slide-number: true
    show-slide-number: print
  #multiplex: true
    logo: ORU-logo_LR.jpg
    footer: "Computing and Mathematics Department. (Dr. V.)"
    chalkboard: true
    # css: footnote-floating.css
number-sections: true
toc: true
---

## What is a differential equation? {.scrollable}

::: {.callout-tip}
## Definition:

A `differential equation` is an equation relating an unknown function and any of its derivatives or differentials
:::

. . . 

<span style="color: grey;font-size: 0.5em;">*Press "b", and ask class for examples* </span>


## Example: Growth and Decay {.scrollable}

::: {.callout-tip collapse="true"}

## Modeling: Growth and Decay

Model $$\boxed{y'=ay}$$\
Initial Condition (<span style="color: red;">i.c.</span>): $$y(0)=y_0$$
:::

. . . 

<span style="color: grey;font-size: 0.5em;">*Press "b", Let's solve this together...* </span>

. . . 

::: {.callout-caution collapse="true"}
## Solution: Growth and Decay


```{=tex}
\begin{align}
y'&=ay \\
\frac{dy}{dt} &= a y \quad{\text{(Separation of variables)}}\\
\frac{dy}{y} &= a dt \\
\int \frac{1}{y}\,dy &=\int a \,dt\\
\ln y &= at + c\\
y&=e^{at+c}\\
y&=e^{at}e^{c}\\
y(0)&=y_0\quad\text{(Initial condition)}\\
\therefore y&=y_0e^{at}
\end{align}
```

:::

. . . 

**Question:** How the value of $a$ affect the solution?

<span style="color: grey;font-size: 0.5em;">*Press "b", sketch the "solutions" ...* </span>


. . .

::: {.callout-caution collapse="true"}
## Viz: Growth and Decay

We can use `R` to do the plots:

```{r, echo=TRUE}
#| echo: fenced
y0 <- 100
t <- seq(0, 10, length.out = 100)
a1 <- 1
a2 <- -1
y1 <- y0*exp(a1*t)
y2 <- y0*exp(a2*t)
plot(t,y1)
legend("center", legend= "a>0")
plot(t,y2)
legend("center", legend= "a<0")
```

:::

. . . 

**Note:** You want to see a video example? Click here: @sec-ref

## Example: Newton's law of cooling {.scrollable}

::: {.callout-tip}

## Newton's law of cooling

$$\boxed{\frac{dT}{dt}=-k(T-T_{env})}\quad T(0)=T_0;\quad k>0$$
:::
. . . 

<span style="color: grey;font-size: 0.5em;">*Press "b", How we solve this? ...* </span>

. . . 

::: {.callout-caution collapse="true"}

## Solution: Newton's law of cooling

```{=tex}
\begin{align}
\int \frac{1}{T-T_{env}}dT&=\int -kdt\\
\ln (T-T_{env}) &= -kt+C\\
T-T_{env} &= e^{-kt+C}\\
T-T_{env} &= e^{-kt}e^{C}\quad\Rightarrow\text{(Initial conditions)}\\
T-T_{env} &= e^{-kt}(T_0-T_{env})\\
T&=T_{env}+(T_0-T_{env})e^{-kt}
\end{align}
```

:::

. . . 

<!-- ## Sol. Curves: Newton's law of cooling {.scrollable} -->

::: {.callout-caution collapse="true"}
## Solution Curves: Newton's law of cooling

```{r, echo=TRUE}
#| echo: fenced
T0 <- 1000
Tenv <- 300
k <- 1
t <- seq(0, 10, length.out = 100)
y <- Tenv + (T0-Tenv)*exp(-k*t)
plot(t,y, ylim = c(0,1000))
legend("center", legend="Converging to Tenv")
```
:::

## Newton's Second Law {.scrollable}

::: {.callout-tip}
## Newton's Second Law

"Force is equal to the rate of change of momentum" $$\boxed{F=\frac{d}{dt}(mv)}$$ If `mass` is constant then $F=m\frac{dv}{dt}=ma$
:::

. . . 

::: {.callout-caution collapse="true"}
## Solution
\begin{align}
\frac{dx^2}{dt^2}&=\frac{F(t)}{m}\\
\text{Integrating:}\\
v(t)&=v_0+\frac{1}{m}\int_0^t F(\tau)d\tau
\end{align}
:::

## Definition to be a solution of diffyQ {.scrollable}

::: {.callout-tip}
## Definition

A function $y(x)$ is said to be a `solution` to a differential equation $\Leftrightarrow$ an identity results when the corresponding values for $y(x)$ and its derivatives are substituted into the equation.
:::

. . . 

::: {.callout-caution collapse="true"}
## Example

Verify that $y=e^x+x+7$ is a solution to $$\frac{dy}{dx}-y+x=-6$$
:::

. . .

<span style="color: grey;font-size: 0.5em;">*Press "c", Let's check if its solution ...* </span>

. . .

::: {.callout-caution collapse="true"}
## Solution:

```{=tex}
\begin{align}
\frac{dy}{dx}-y+x&=-6\quad\text{(expanding the left side:)}\\
\frac{d}{dx}(e^x+x+7)-(e^x+x+7)+x&=-6\\
e^x+1-e^x-x-7+x&=-6\\
-6&=-6
\end{align}
```
😁
:::

## Explicit and Implicit form of a solution

::: {.callout-tip}
## Definition

1.  A solution is in an `explicit form` $\Leftrightarrow y=f(x)$
2.  A solution is in an `implicit form` $\Leftrightarrow f(x,y)=0$
:::

## Is the sol. of a diffyQ always unique? {.scrollable}

::: {.callout-warning collapse="true"}
## **Q:** Is the solution of a diffyQ always unique?

:::

. . . 

A: No

. . .

::: {.callout-caution collapse="true"}
## Example ^[<https://en.wikipedia.org/wiki/Harmonic_oscillator>]

$$\frac{d^2y}{dx^2}+y=0$$
:::

. . . 

::: {.callout-caution collapse="true"}
## Solution:

:::

. . .

By inspection we observe that two different functions satisfy:

```{=tex}
\begin{align}
y_1&=\cos x\\
y_2&=\sin x\\
\end{align}
```

. . .

Therefore the most `General Solution` is a Linear Combination of the particular solutions:

$$ \boxed{y = C_1\cos x + C_2\sin x }$$

## Does every diffyQ have a sol$^\underline{n}$ ?{.scrollable}

::: {.callout-warning}
## **Q:** Does every diffyQ have a sol$^\underline{n}$ ?

:::

. . .

A: No

. . .

::: {.callout-caution collapse="true"}
## Example

$$\left(\frac{dy}{dx}\right)^2+1=0$$
:::

. . .

Any solution to this eq$^\underline{n}$ would have an `imaginary` slope!

Thus no-real solution!^[interested in imaginary? $\Rightarrow$ PHY:455<https://syllabi.oru.edu/?id=28002>]


## Have questions ?

Schedule Office Hours with your Instructor Here!

<!-- Calendly inline widget begin -->
<div class="calendly-inline-widget" data-url="https://calendly.com/evalderrama/officehours" style="min-width:320px;height:630px;"></div>
<script type="text/javascript" src="https://assets.calendly.com/assets/external/widget.js" async></script>
<!-- Calendly inline widget end -->


## References {#sec-ref}

1.  Exponential growth and decay. <https://en.wikipedia.org/wiki/Exponential_growth>
2.  Newton law of cooling. <https://en.wikipedia.org/wiki/Newton%27s_law_of_cooling>
3.  Harmonic oscillator. <https://en.wikipedia.org/wiki/Harmonic_oscillator>
4.  Khan: Modeling population with simple differential equation. <https://youtu.be/IYFkXWlgC_w?feature=shared>
<!-- 3.  Quarto revealjs. <https://quarto.org/docs/presentations/revealjs/themes.html> -->
<!-- 4.  Quarto callouts. <https://quarto.org/docs/authoring/callouts.html> -->
